\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1.0in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{physics}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{amsthm}
\usepackage{pgfplots}

\pgfplotsset{compat=1.16}



\renewcommand{\thesubsection}{\thesection\Alph{subsection}}
\renewcommand\qedsymbol{\square}



\title{Classical Mechanics  PS3}
\author{Joe Crowley}
\date{October 2020}

\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{1pt}

\fancyhf{}
\rhead{
Joe Crowley \\
Physics 205 \\
Problem Set 3\\
}
\rfoot{Page \thepage}

\begin{document}  

\section{GPS 3.11}
\textit{Two particles move about each other in circular orbits under the influence of gravitational forces, with a period $\tau$. Their motion is suddenly stopped at a given instant of time, and they are then released and allowed to fall into each other. Prove that they collide after a time $\tau / 4 \sqrt{2}$.}


\begin{equation*}
    L=\frac{1}{2} \mu\left(r^{2}+r^{2} \dot{\theta}^{2}\right)+\frac{k}{r}
\end{equation*}
\begin{align*}
    \frac{d}{d t}\left(\frac{\partial L}{\partial \dot{r}}\right)-\frac{\partial L}{\partial r}&=0\\
    \mu \ddot{r}-\mu r \dot{\theta}^{2}+\frac{k}{r^{2}}&=0\\
    &\\
    \frac{d}{d t}\left(\frac{\partial L}{\partial \dot{\theta}}\right)-\frac{\partial L}{\partial \theta}&=0\\
    \frac{\partial L}{\partial \dot{\theta}}&=\operatorname{const}\\
    \operatorname{mr}^{2} \dot{\theta}=\ell
\end{align*}
Suddenly, $\dot{\theta}=0$. 

\begin{align*}
    \ddot{r}&=-\frac{k}{\mu r^{2}}\\
    \ddot{r} \dot{r}&=-\frac{k}{\mu r^{2}} \dot{r}\\
    \frac{1}{2} r^{2}&=\frac{k}{\mu r}+C_{1}\\
    \dot{r}^{2}&=\frac{2 k}{\mu r}+C_1
\end{align*}

Using initial conditions $r = r_0, \dot r = 0$, $C=-\frac{2 K}{\mu r_{0}}$.

\begin{align*}
    \dot{r}^{2}&=\frac{2 k}{\mu}\left(\frac{1}{r}-\frac{1}{r_{0}}\right)\\
    r&=\sqrt{\frac{2 k}{\mu}}\left(\frac{1}{r}-\frac{1}{r_{0}}\right)^{\frac{1}{2}}\\
\end{align*}

\begin{align*}
    \sqrt{\frac{\mu}{2 k}}\left(\frac{1}{r}-\frac{1}{r_{0}}\right)^{-\frac{1}{2}} d r&=dt\\
    -\sqrt{\frac{\mu}{2 k}}\left(r_{0} r \sqrt{\frac{1}{r}-\frac{1}{r_{0}}}-r^{\frac{3}{2}} \tan ^{-1}\left(\sqrt{r_0\left(\frac{1}{r}-\frac{1}{r_{0}}\right)}\right)\right)\Big|_{r_{0}} ^{0}&=\Delta t\\
    \frac{a^{\frac{3}{2}} \pi \sqrt{\frac{\mu}{k}}}{2 \sqrt{2}}&=\Delta t
\end{align*}

Using $\tau=2 \pi a^{\frac{3}{2}} \sqrt{\frac{\mu}{k}}$,

$$\boxed{\Delta t=\frac{\tau}{4 \sqrt{2}}}$$

\newpage



\section{GPS 3.12}
\textit{Suppose that there are long-range interactions between atoms in a gas in the form of central forces derivable from a potential
$$
U(r)=\frac{k}{r^{m}}
$$
where $r$ is the distance between any pair of atoms and $m$ is a positive integer. Assume further that relative to any given atom the other atoms are distributed in space such that their volume density is given by the Boltzmann factor:
$$
\rho(r)=\frac{N}{V} e^{-U(r) / k T}
$$
where $N$ is the total number of atoms in a volume $V$. Find the addition to the virial of Clausius resulting from these forces between pairs of atoms, and compute the resulting correction to Boyle's law. Take $N$ so large that sums may be replaced by integrals. While closed results can be found for any positive $m,$ if desired, the mathematics can be simplified by taking $m=+1$}
\newpage


\section{GPS 3.19}
\textit{A particle moves in a force field described by
$$
F(r)=-\frac{k}{r^{2}} \exp \left(-\frac{r}{a}\right)
$$
where $k$ and $a$ are positive.}


\subsection{}
\textit{Write the equations of motion and reduce them to the equivalent one-dimensional problem. Use the effective potential to discuss the qualitative nature of the orbits for different values of the energy and the angular momentum.}

\begin{align*}
    L &= T - v \\
     &= \frac{1}{2} m\left(\dot{r}^{2}+r^{2} \dot{\theta}^{2}\right)+\int F \cdot d r\\
\end{align*}
Since $L$ has no explicit time dependence, and the coordinates are natural, the hamiltonian is conserved and is the total energy. 

\begin{align*}
    E&=\frac{1}{2} m\left(r^{2}+r^{2} \dot{\theta}^{2}\right)-\int F \cdot d r\\
     &\frac{d}{d t} E=m\left(\ddot{r} \ddot{r}+r \dot{r} \dot{\theta}^{2}+r^{2} \dot{\theta} \ddot{\theta}\right)-F \dot{r}\\
\end{align*}

From the Euler-Lagrange equation for $\theta$, 

\begin{align*}
    \dot \theta &= \frac{l}{m r^2}\\
    \ddot \theta &= -\frac{2l}{m r^3}\dot r\\
\end{align*}

\begin{align*}
    \dot{r} \ddot{r}&=\frac{F}{m} \dot{r}-r \dot{r} \frac{l^{2}}{m^{2} r^{4}}-r^2 \frac{l}{m r^{2}}\left(-\frac{2 l}{m r^{3}}\right) \dot{r}\\
    \ddot{r}&=\frac{F}{m}+\frac{l^{2}}{m^{2} r^{3}}\\
\end{align*}


$$\boxed{\ddot{r}=\frac{l^{2}}{m^{2} r^{3}}-\frac{k}{r} e^{-\frac{r}{a}}}$$

Using the effective potential, orbits are $\approx$ elliptical when $\partial_r V_\mathrm{eff} = 0$.

\begin{align*}
    V_{\text {eff }}&=-\int F d r+\frac{l^{2}}{2 m r^{2}}\\
    &\\
    \frac{d V_\mathrm{eff}}{d r}&=\frac{d V}{d r}-\frac{l^{2}}{m r^{3}}\\
    0&=-F-\frac{l^{2}}{m r^{3}}\\
    \frac{k}{r^{2}} e^{-\frac{r}{a}}& =\frac{\ell^{2}}{m r^{3}}\\
    \frac{k m}{l^{2}} e^{-\frac{r}{a}}& =\frac{1}{r}\\
\end{align*}

For some value of $\frac{k m}{l^{2}}$, the LHS and RHS will have the same derivative, 

\begin{align*}
    -\frac{k m}{a l^{2}} e^{-\frac{r}{a}}&=-\frac{1}{r^{2}}\\
    \frac{1}{a r}&=\frac{1}{r^{2}}\\
\end{align*}

$r=a$ is a saddle point. Closed orbits exist for some values of $r$ when $V_\mathrm{eff} < E <0, V_\mathrm{eff}>E>0$, and circular orbits for $V_\mathrm{eff} = E$.

\subsection{}
\textit{Show that if the orbit is nearly circular, the apsides will advance approximately by $\pi \rho / a$ per revolution, where $\rho$ is the radius of the circular orbit.}
$$u = u_0 + \alpha \cos \beta \theta$$

\begin{align*}
    \frac{1}{r} &= \frac{1}{\rho} + \alpha \cos \beta \theta\\
    &\\ 
    \beta^2 &= 3 - \frac{r}{F}\frac{d F}{dr}\Big|_{r = \rho}\\
    &= 1 - \frac{\rho}{a}\\
    \beta &\approx 1 - \frac{\rho}{2a}\\
    &\\
    \theta &= \frac{2 \pi}{\beta}\\
    &\approx 2 \pi (1 +  \frac{\rho}{2a})\\
\end{align*}
This shows that for every $2\pi$ radians, the apsides will include $\frac{\pi\rho}{a}$ in excess. 



\newpage


\section{GPS 3.21}
\textit{Show that the motion of a particle in the potential field}
$$
V(r)=-\frac{k}{r}+\frac{h}{r^{2}}
$$
\textit{is the same as that of the motion under the Kepler potential alone when expressed in terms of a coordinate system rotating or precessing around the center of force.}

\textit{For negative total energy, show that if the additional potential term is very small compared to the Kepler potential, then the angular speed of precession of the elliptical orbit is}
$$
\dot{\Omega}=\frac{2 \pi m h}{l^{2} \tau}
$$
\textit{The perihelion of Mercury is observed to precess (after correction for known planetary perturbations) at the rate of about $40^{\prime \prime}$ of arc per century. Show that this precession could be accounted for classically if the dimensionless quantity}
$$
\eta=\frac{h}{k a}
$$
\textit{(which is a measure of the perturbing inverse-square potential relative to the gravitational potential) were as small as $7 \times 10^{-8}$. (The eccentricity of Mercury's orbit is $0.206,$ and its period is 0.24 year.)}

\begin{align*}
    \frac{d^{2} u}{d \theta^{2}}+u&=-\frac{m}{l^{2}} \frac{d}{d u}\left(-k u+h u^{2}\right)\\
    &=\frac{m k}{l^{2}}-\frac{2 h m}{l^{2}} u\\
\end{align*}
\begin{align*}
    \frac{d^{2} u}{d \theta^{2}}+(1+\frac{2 h m}{l^{2}}) u&=\frac{m k}{l^{2}}\
\end{align*}

Let $\lambda^2  = 1+\frac{2 h m}{l^{2}}$.
$$
    \frac{d^{2}}{d \theta^{2}}\left(u-\frac{m k}{\lambda^{2} l^{2}}\right)+\lambda^{2}\left(u-\frac{m k}{\lambda^{2} l^{2}}\right)=0
$$
Which has real solutions 
$$ u = \frac{m k}{\lambda^{2} l^{2}} + A \cos \lambda\theta + B \sin \lambda \theta$$.


At perihelion, 
\begin{align*}
    \frac{1}{r_{\min }}-\frac{m k}{\lambda^{2} l^{2}}&=A\\
    \lambda B \cos \lambda \theta &=0
\end{align*}

$$
    u = \frac{m k}{\lambda^{2} l^{2}} ( 1 - \cos \lambda \theta) + \frac{\cos \lambda \theta}{r_\mathrm{min}}
$$
$$
r=\frac{\lambda^{2} l^{2}}{m k} \frac{1}{1+e \cos \lambda \theta}
$$

\begin{align*}
    \theta&=\frac{2 \pi}{\lambda}\\
    &=\frac{2 \pi}{\sqrt{1+\frac{2 h m}{l^{2}}}}\\
    &=2 \pi\left(1+\frac{2 h m}{l^{2}}\right)^{-\frac{1}{2}}\\
    &=2 \pi-\frac{2 \pi m h}{l^{2}}
\end{align*}

this corresponds to a perihelion precession of $\frac{2 \pi m h}{l^{2}}$ every period. 

\begin{align*}
    \dot\Omega &= \frac{2 \pi m h}{\tau l^{2}}\\
    &=\frac{2 \pi}{\tau} \frac{1}{1-e^{2}} \frac{h}{k a}\\
    &=\frac{2 \pi}{\tau} \frac{\eta}{1-e^{2}} 
\end{align*}

\begin{align*}
    \eta &= \frac{\left(1-e^{2}\right) \tau \dot{\Omega}}{2 \pi}\\
    &= \frac{\left(1-0.206^{2}\right) (0.24 \mathrm{years}) (\frac{40^{\prime\prime}}{\mathrm{century}})}{2 \pi}\\
    &= 7.09307 \times 10^{-8}
\end{align*}

\newpage

\section{GPS 3.22}
\texit{The additional term in the potential behaving as $r^{-2}$ in Exercise 21 looks very much like the centrifugal barrier term in the equivalent one-dimensional potential. Why is it then that the additional force term causes a precession of the orbit, while an addition to the barrier, through a change in $l,$ does not?}
the additional force term causes a precession in the orbit because it acts as a centrifugal term, however the term is more significant for larger l because the term is $\propto \frac{1}{l^2}$. 


\end{document}
